Final project

### 2025-04-17 code/full_code/full_code.qmd
Few minor changes

## General Background Information

Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease. It’s also what most people mean when they use the general term “heart disease”. It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart. This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath. As the leading cause of death in the U.S. and globally, CAD claimed 375,500 lives in the U.S. in 2021 and affects over 18 million American adults.

## Description of data and data source

This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses. Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).

This heart disease dataset is acquired from one o f the multispecialty hospitals in India. This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type). This dataset was obtained from mendeley data where it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

## Questions/Hypotheses to be addressed

Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820

## Methods

In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present. The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).

We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values. Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression. Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.

To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups. Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar. Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.

In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods. These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE). By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.

## **Conclusion**

Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present. Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups. In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance. Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup. Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease. Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.

## Data aquisition

This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

@@

```{r}
# Calculate summary statistics for selected variables
summary_stats <- cardio_df %>% 
  summarise(
    # Age statistics
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    # Resting Blood Pressure statistics
    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    # Serum Cholesterol statistics (excluding zeros)
    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),

    # Maximum Heart Rate statistics
    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    # Oldpeak statistics (excluding zeros)
    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
  )

# Print summary statistics
print(summary_stats)

summary_df = skimr::skim(summary_stats)
print(summary_stats)
# save to file
summarytable_file = here("results", "tables", "summarytable1.rds")
saveRDS(summary_df, file = summarytable_file)

@@

The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup. For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels. Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.

@@

In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero. However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own. When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance. These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.

@@

# save fit results table  
table_file1 = here("results", "tables", "resulttable3.rds")
saveRDS(or_results, file = table_file1)

@@

The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87). For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22). The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82). However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.

@@

# save fit results table  
table_file1 = here("results", "tables", "resulttable4.rds")
saveRDS(rr_results, file = table_file1)
```

The baseline relative risk is about 0.23 (95% CI: 0.08–0.46). For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35). The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.

@@

```{r}
roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
auc_value <- auc(roc_curve)
print(paste("AUC:", auc_value))
```

The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing. It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.

@@

The adjusted logistic regression model produces unstable estimates. The intercept's odds ratio is about 0.07, indicating very low baseline odds. The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant. The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals. These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable. This interpretation is derived solely from your output.

@@

The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average. This indicates limited accuracy for distinguishing between individuals with and without heart disease.

@@

The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average. This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.

@@

@@

The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half. This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.

@@

# Plot LASSO tuning results
ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 2, color = "firebrick") +
  scale_x_log10() +
  labs(
    x = "Penalty (log scale)",
    y = "RMSE",
    title = "LASSO Tuning (No Cross-validation)",
    subtitle = "Using apparent resampling"
  ) +
  theme_minimal()
```

The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties. This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.

@@

# Plot RF tuning results
ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Random Forest Tuning (No Cross-validation)",
    x = "mtry",
    y = "min_n",
    fill = "RMSE"
  ) +
  theme_minimal()
```

The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739. This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used. In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.

@@

Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases. This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.

@@

Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry). In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values. This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

>>>>

@@

```{r}
# Calculate summary statistics for selected variables
summary_stats <- cardio_df %>% 
  summarise(
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    SerumCholestrol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),

    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
  )

# Reshape into long format, then wide table
summary_table_formatted <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Variable, values_from = value)

# View formatted table
summary_table_formatted

# save to file
summarytable_file = here("results", "tables", "table1_summary_statistics.rds")
saveRDS(summary_table_formatted, file = summarytable_file)

@@

@@

@@

# save fit results table  
table_file2 = here("results", "tables", "table2_odds_ratios.rds")
saveRDS(or_results, file = table_file2)

@@

@@

# save fit results table  
table_file3 = here("results", "tables", "table3_relative_risks.rds")
saveRDS(rr_results, file = table_file3)
```

@@

```{r}
roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
auc_value <- auc(roc_curve)
print(paste("AUC:", auc_value))

# Save ROC plot
roc_fig <- recordPlot()
saveRDS(roc_fig, file = here("results", "figures", "figure1_roc_logistic_model.rds"))

@@

@@

@@

@@

# Save RMSEs from all models to a table
rmse_df <- tibble(
  Model = c("Linear", "LASSO", "Random Forest"),
  RMSE = c(
    rmse(cleaned_df, truth = target_binary, estimate = pred_lm)$.estimate,
    rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)$.estimate,
    rmse(cleaned_df, truth = target_binary, estimate = pred_rf)$.estimate
  )
)
saveRDS(rmse_df, file = here("results", "tables", "table5_model_rmse_comparison.rds"))

@@

@@

# Identify best LASSO penalty
best_lasso <- lasso_tune_df %>% filter(.estimate == min(.estimate))
print("Best LASSO Penalty (No CV):")
print(best_lasso)

# Plot LASSO tuning results
ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 2, color = "firebrick") +
  scale_x_log10() +
  labs(
    x = "Penalty (log scale)",
    y = "RMSE",
    title = "LASSO Tuning (No Cross-validation)",
    subtitle = "Using apparent resampling"
  ) +
  theme_minimal()

# Save LASSO tuning plot
lasso_tune_plot <- last_plot()
ggsave(filename = here("results", "figures", "figure2_lasso_tuning_plot.png"), plot = lasso_tune_plot)
```

@@

# Identify best RF tuning parameters
best_rf <- rf_tune_df %>% filter(.estimate == min(.estimate))
print("Best Random Forest Parameters (No CV):")
print(best_rf)

# Plot RF tuning results
ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Random Forest Tuning (No Cross-validation)",
    x = "mtry",
    y = "min_n",
    fill = "RMSE"
  ) +
  theme_minimal()

# Save Random Forest tuning heatmap
rf_tune_plot <- last_plot()
ggsave(filename = here("results", "figures", "figure3_rf_tuning_heatmap.png"), plot = rf_tune_plot)

@@

@@
